SUPPORTING INFORMATION \\ Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-07-15 cover_letter/response_letter.tex

\end{letter}

%----------------------------------------------------------------------------------------

\clearpage
\newpage

\noteg{My notes and questions are all in green. The rest of the color scheme is the same as the example you sent me. For most comments I have attempted some change in this doc but have not changed anything major in the actual manuscript or tracked version (manuscript\_marked\_up.tex) yet. In some places I have proposed a couple options but did not want to delve to deep in case I'm going down the wrong path.}

@@

\textbf{Author Reply:}   We thank the reviewer for their close reading of the manuscript and are pleased they find it significant to JACS readers.

\textcolor{blue}{Major point.
1)      Many of the core conclusions about the hybridization and dehybridzation mechanisms are based on the designation of the relevant metastable states as H, 5S2, 3S2, 5S4, F4, and D. While the authors do an excellent job of demonstrating that 2-6 state models provide an accurate description of the dynamics, in the current form, the paper does not provide sufficient evidence to support their chosen representative microstates and associated nomenclature (the only evidence being the ‘representative’ structures shown in Figures 2A, C). A given macrostate represents an ensemble of microstates and it should be demonstrated that a significant fraction of the microstates that have a high probability of belonging to a particular macrostate share the same or similar base-pairing pattern. This would better justify how the authors chose to describe these states as 5S2, 3S2, etc., and provide the reader with more confidence that the mechanisms follow from the simulations. At the least, the authors should describe how the representative microstates (depicted in Figure 2A, C) were obtained.
}

\textbf{Author Reply:}   We acknowledge that the designation of states is crucial to the core conclusions of the manuscript and that some information regarding state definitions and representative structures was unclear.

\noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering. I've tried to address both aspects below and can add more detail the first figure or choose a different kind of representation.}

When defining macrostates, we both visually inspected configurations and compared distributions along a variety of structural features.  During the analysis process, we have found that two distance features in particular allow for interpretable visualization of each of the seven macrostate. Specifically, these features correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5$\prime$ and 3$\prime$ directions. These features highlight not only shifted states in both directions, but also a distinct frayed regions between the H and D wells. We have visualized these features below (Figure \ref{fig:macrostate_colors_FES}), and use the PyEMMA free energy function to plot the distribution of each macrostate/sequence on the low dimensional feature representation. Each color map corresponds to a different macrostate according to the legend, and lighter regions indicate a higher probability at equilibrium (lower free energy). We agree with the reviewer that it is important for readers to understand the physical definitions and distributions of each macrostate and its associated microstates, and we have added this figure to the supplemental information.

@@

%In order to demonstrate that configuration in a given macrostates are well-defined by the representations in figure 2, we show the distribution of several structural metrics for each macrostate and sequence combination. These metrics are a collection of mean intermolecular distance that we have found work well to differentiate between states.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=150mm]{cover_letter/revision_figures/macrostate_colors_FES.png}
        \caption{Free energy surface of each macrostate and sequence combination. The x and y axes correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5$\prime$ and 3$\prime$ directions, respectively. These distances are visualized by the corresponding red arrow representations. Color maps define both the macrostate identity and relative free energy of a region within that macrostate, lighter regions correspond to lower free energy.}
        \label{fig:macrostate_colors_FES}
	\end{center}
\end{figure}

With regard the molecular renderings shown in figures 2A and 2C, these were chosen from microstates with the highest probability of being assigned to their corresponding macrostates. Accordingly, they should correspond to structures  that are most representative of that macrostate as whole. As the reviewer points out, it is unclear how these selected microstates were defined in the main text, and we include the following addition to section 3.2 to make this point explicit:

``We present these seven macrostates in Fig. 2a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. \rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole. A more detailed representation of macrostate membership is shown in supplemental figure \noteg{add supplemental reference here}''}

@@

\textbf{Author Reply:}   Five-fold cross-validation was conducted by randomly partitioning 40 independent trajectories into two sets of 20. In this way, all data is included in each fold but with distinct test/train sets. This approach prevents rare events from being left out of small validation sets and was used in our previous study, where one long trajectory was broken up into 100 \citep{Sidky2019High-ResolutionVAMPnets} smaller trajectories for training. We have  clarified this in the text by adding the following sentence to section 2.1:

We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits \rood{such that random combinations of 20 trajectories were partitioned into each set}.\citep{Sidky2019High-ResolutionVAMPnets}

\textcolor{blue}{2)      For the validation of embedding dimensions (P. 11 L. 3), it is hard to judge the presence of a kink at 5D in the AT-all sequence since only a maximum of 6D is shown in Figure S1. Could a few additional dimensions be provided?}

\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences. As suggested by the reviewer, we added two additional modes to the cross-validation procedure showing this kink after five modes and have updated the AT-all subplot of Figure S1 with the panel shown in Figure \ref{fig:AT-all_crossval}.

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=75mm]{cover_letter/revision_figures/AT-all_added_modes.png}
        \caption{Cross-validation for the AT-all sequence with two additional modes to show the presence of a kink in the VAMP-2 score after the fifth mode.}
        \label{fig:AT-all_crossval}
	\end{center}
\end{figure}

\textcolor{blue}{3)      The values associated with the “slow” and “fast” rates appear to be inverse quantities, but this is not mentioned in the text (P. 14 L. 44).}

\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 14. We agree that the sentence on (P. 14 L. 44) in unclear in its current form and have made the following change:

@@

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Figure 1 has been updated to include units in s$^{-1}$. In accordance with suggestions by reviewer \#3 we have also displayed the slow responses on a y-log axis, and used the finite size effect correction mentioned by reviewer \#2 to improve the fit (Figure \ref{fig:conc_correction}). \noteg{Should we include the figure here instead? Thought it made more sense in the R2 section since we are addressing a more substantial change}

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all Figure S4 plots to include goodness-of-fit values determined by the $R^2$ values of log fits (Figure \ref{fig:goodness-of-fit}). We have also re-plotted these on the log axis for easier comparisons between the simulated data and fits. Although all fits are strong, it does appear that GC-core may be slightly worse than the others sequences. As the reviewer suggests, this may be a contributing factor to the weaker experimental agreement for GC-core as well, however it is difficult for us to make that claim definitely given the data. As we note in the original text, this discrepancy may also be a result of its high propensity to fray. \noteg{Not sure if we really want to go into this or update the main text?}

@@

\textcolor{blue}{6)       The authors chose to compute the temperature jump relaxation rates from fits to quantities obtained directly from the MD simulation. However, it is not clear why the authors did not also compute the relaxation of observables from estimated Markov state models following (P. N. A. S. USA. 2011;108(12):4822-4827). This approach has also been previously used to compare kinetics obtained from Markov state models with other spectroscopic temperature-jump based experiments in DNA (J. Phys. Chem. B. 2019;123(10):2291-2304.) and RNA (J. Chem. Theory Comput. 2017;108(2):926–934) oligomers which should be mentioned. If possible, it would be interesting to compare or discuss the differences between their approach and these previous approaches.}

\textbf{Author Reply:}   The reviewer raises an important point regarding previous methodologies that have been developed to directly recapitulate experimental relaxation rates from MD-derived MSMs.

\noteg{This paragraph may not be necessary}
We have since experimented with some of these techniques using built-in PyEMMA functions, and although the results are interesting, they do not provide the most informative analysis of these particular dynamics or a strong experimental connection. We believe this is the case for the following reasons: i) shifting dynamics occur on timescales approaching that of dissociation and ii) shifting populations are small in amplitude relative to fraying and full dissociation and iii) a shifting signal is difficult to interpret experimentally without a more advanced technique such a as isotope labeling. \noteg{check with Brennan here} The combined result of these effects is that shifting dynamics are thought to make small contributions blended with the slow dissociation response. We support this hypothesis with data shown in Figure \ref{fig:responses_all}) \noteg{not sure how much to go into here}.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=150mm]{cover_letter/revision_figures/responses_all.png}
        \caption{a) Experimental rate representations for each sequence. Both the 1600 cm$-1$ corresponding to A,T dynamics and 1660$-1$ corresponding to G,T dynamics are shown. b) Simulated relaxation using MSM built from each sequence. An A:T and G:C response are calculated based on the distance between available WC pairs. c) Change in response when factoring in shifting motions. Because shifting is low in amplitude and occurs on timescales approaching full dissociation, the response is not expected to be clearly discernible experimentally.  \noteg{need more description here, but not sure how much of this to include} }
        \label{fig:responses_all}
	\end{center}
\end{figure}

@@

% Can also describe fraying issue
% The appraoch in PNAS assumes that the dynamical observable can be captured by a markovian lag time
The second problem we encountered with the MSM relaxation approach is that it does not adequately describe fast-termini fraying. As discussed in section 3.1, the experimental fast responses is thought to corresponds to single base fraying of (mostly A:T) termini. This is a process that occurs even faster in simulation as shown by our derived acceleration factor. We found that although fraying does occur for each sequence, this process cannot be captured by a Markovian lag time. On the other hand, multi-base fraying observed in GC-core is slower and corresponds to the second leading slow mode for that sequence. As shown in Figure \ref{fig:responses_all}b, we only see a distinct fast response for GC-core despite the fact that termini fraying is occurring in each sequence (as demonstrated by Figure 1 in the main text).

In addition to the issues discussed above, the main reason that we extracted responses directly from MD simulations was that effects of coarse graining, concentration, salt, etc. were unknown and could not be accounted for at a single temperature. For the slow response in particular,  timescales have an exponential dependence on temperature which could not be sufficiently standardized between experiment and simulation. Furthermore, generating equilibrium simulation data over a range of temperatures for each sequence was computationally prohibitive, prompting our decision to run an ensemble of shorter simulations at each temperature.

The additional references cited by the reviewer highlight important related work and should be included \citep{Noe2011DynamicalExperiments, Pinamonti2017, Remington2019MolecularKinetics}. Although we discussed our work's connection to (J. Chem. Theory Comput. 2017;108(2):926–934)\citep{Pinamonti2017} in the introduction, we did not originally reference it with respect to relaxation rate simulations. We do believe that our work is quite distinct from these previous studies, given that much smaller RNA oligonucleotide systems were investigated. Furthermore, the RNA studies were either lacking experimental data or compared against older data which suffered some experimental flaws. This highlights the difficulty in making strong experimental comparions for oligonucleotide systems using MSM relaxation techniques alone.

@@

We have added the following text to section 3.1 make some of these difficulties more explicit and qualify our approach:

\noteg{Not totally sure where do put this, maybe end of first paragraph of 3.1 or at the end of that section?}

\rood{"Previous studies have used a dynamical fingerprinting approach \citep{Noe2011DynamicalExperiments} to compute rates from MSMs. These techniques were used to study base stacking dynamics in 2-4 base RNA oligomers \citep{Pinamonti2017, Remington2019MolecularKinetics} and some attempts were made to compare against experiment. However, Limitations in all-atom force fields and potential experimental artefacts have made strong comparisons difficult to achieve for oligonucleotide systems \citep{Pinamonti2017}. We tried applying these techniques with our SRV-MSMs as well, but found that i) termini fraying timescales were too fast for a Markovian lag time and ii) coarse-graining effects could not be properly standardized between experiment and simulations at a single temperature. This motivated us to directly analyze responses from simulation data over a range of temperatures."}

@@

\begin{shaded}
\textbf{Response to Reviewer \# 2}
\end{shaded}

\textcolor{blue}{This paper reports simulation results on the hybridization and dehybridization mechanisms near to the melting temperature of a number of short DNA sequences, and accompanying temperature jump IR experiments. The sequences are chosen to have a repeating AT motif that potentially allows relatively stable out-of register duplex states, with a focus being on how these states are disrupted, by the presence of GC base pairs. The authors show that the introduction of just 2 appropriately-positioned base pairs can change the kinetics to a simple two-state behaviour from one in which multiple states contribute.}

\textbf{Author Reply:}   We thank the reviewer for their comments. \noteg{need to say anything here?}

@@

\textcolor{blue}{I found this an interesting study that nicely illustrated some of the potential effects of sequence on (de)hybridization mechanisms. However, I found the results unsurprising, and these types of effects have been previously explored in the literature. Although I can't therefore recommend publication in JACS, I think it would be very appropriate for one of the more specialized ACS journals.}

\textbf{Author Reply:}  We are glad that the reviewer finds our results interesting, and we acknowledge that there is precedent in the literature for the states and mechanisms that we discuss.  Indeed, although similar dynamics have been identified previously, the transition probabilities between states and sequence-dependent affect on these transitions has not been identified and quantified. As suggested by reviewer \#2 and later by reviewer \#3, we have updated several introductory sentences to specify that the transition network between states reflects is significance of this work rather than the discovery of the states or mechanisms themselves

@@

\noteg{Add a small methods paragraph to describe the temp-dependent fits? I think this would also help address reviewer \#3 confusions about the various temps used }

We have since incorporated the finite size correction for homodimers reported in  https://doi.org/10.1088/0953-8984/22/10/104102 \citep{Ouldridge2010ExtractingSimulations} when computing the slow dissociation fits. We find that the results show even better agreement with experiment, motivating us to update figure 1 with these data as shown below \ref{fig:conc_correction}. In addition, we have added the units and log scale changes proposed by reviewers \#1 and \#3, respectively.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_4K_shift.png}
        \caption{Updated figure 1 of the main text. We made three changes: i) adding units to both plots, ii) accounting for finite size effects when computing the slow response, and iii) displaying the slow response on a log axis. These were proposed by reviewers \#1, \#2, and \#3 respectively}
        \label{fig:conc_correction}
	\end{center}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{cover_letter/revision_figures/both_response_log_no_correction_4K_shift.png}
        \caption{Updated figure 1 of the main text. We added units to both plots and displayed the slow response on a log axis as proposed by reviewers \#1 and \#3, respectively \noteg{This is with no correction and the results do not look all the different (homodimer correction is less than hetero). I'm leaning towards just using this given that the temp-dependent simulations are not performed in equlibrium and the correction appears unncessary}}
        \label{fig:conc_no_correction}
	\end{center}
\end{figure}

@@

\textbf{Author Reply:}    This typo has been corrected.

\begin{shaded}
\textbf{Response to Reviewer \# 3}
\end{shaded}

\textcolor{blue}{Jones et al present a combined experimental and computational paper that probes the dynamics of the DNA hybridization/de-hybridization transition. Through a Markov state model analysis of coarse-grained simulation, the authors identify states that contribute to the process, and present experimental evidence to support the role of these states. In particular, repetitive sequences are observed in simulations to have a number of misaligned states that are both appreciably stable and contribute to hybridization and de-hybridization pathways as intermediates. The role of these states is diminished by base pairs that disrupt the repetitive sequence. In the special case of a sequence with a strong core and weaker extremities, metastable "frayed" states are also observed. The presence of these states is assessed experimentally trough the analysis of relaxation timescales observed in T-jump IR spectroscopy; fast relaxation is attributed to fraying equilibria, and a broader fast relaxation to an increased heterogeniety due to the presence of mis-aligned intermediates.}

\textbf{Author Reply:}   We appreciate the reviewer's detailed summary of our work

@@

\textcolor{blue}{The paper is generally well-written, and, I believe, represents a good attempt to tackle this important process. I personally have had a manuscript "desk-rejected" by JACS on the basis that it is "too focussed on DNA biotechnology", and "primarily of interest to RNA and DNA chemists". The same statements could be made about this paper. However, I disagree with the conclusion that such papers are not appropriate for JACS. I would therefore recommend publication, subject to satisfactory responses to the comments below.  }

\textbf{Author Reply:}   We are pleased that the reviewer finds the paper both well-written and of high significance to JACS readers.

\textcolor{blue}{1.   I was frequently confused as to the temperatures used in the experiments and analysis. At one point, it is noted that simulations are performed at melting temperatures. However, simulations are actually performed at a range of temperatures and it isn't always clear when looking at a figure which temperatures were used for simulation, NN model comparison or experiments (see eg. Figs 2 and 3 and 5). I would recommend noting the temperature used to obtain the data in each case.}

\textbf{Author Reply:}   This is a good comment that requires some clarification in the text. We should be more explicit that the MSM data was all acquired at sequence-specific melting temperatures, but the experimental validatoin in 3.1 was performed over a range of temperature in experiment and simulation.

\noteg{we don't really have a methods section on the temp-dependent simulations, maybe this should be added to 2.1.1 address this comment and a question from reviewer \#2? Alternatively we could just add something like this to the intro in addition to another sentence or two?}

"We validate the new computational models of hybridization/dehybridization dynamics developed in this work against new experimental data \rood{over a range of temperatures} and reinterpret our prior experimental observations in light of the new computational understanding."

@@

\textbf{Author Reply:}2.    We appreciate the reviewer pointing out this error in our reported units. The value should be 7 mM, not 7 M as shown in the below calculation. This has been corrected in the manuscript.

### 2021-07-15 cover_letter/response_letter.tex

    \begin{equation}\label{}
	\frac{2 \mathrm{\:oligos}}{\mathrm{box\:volume}} = \frac{2}{7.80\mathrm{\:nm}^3} = 
	\frac{3.32 \times 10^{-24} \mathrm{\:moles}}{4.75 \times 10^{-25} \mathrm{\:m}^3} = 
	6.99 \frac{\mathrm{moles}}{\mathrm{m}^3} \approx 7\mathrm{mM}
	\end{equation}

\textcolor{blue}{3.  The authors repeatedly refer to "slithering" mechanisms for rearrangement. I believe this is unfortunate. In the context of the 3SPN model, the "slithering" term was first introduced in Ref. 49 by Sambriski et al, and describes two helical single strands sliding along each other. As noted by Hinckley et al in Ref 18, that mechanism is largely an artifact of the form of the potential in 3SPN.1. The rearrangement mechanism in 3SPN.2 is more of a "defect diffusion", analogous to the "inchworm mechanism" previously identified in the oxDNA model (ref 52 - note, the author list for this reference is wrong). Given this context, it seems unnecessarily confusing to talk about "slithering".}

\textbf{Author Reply:}    We share the reviewer's concerns with using ``slithering'' to describe a process that would be more appropriately called ``defect diffusion'' or the ``inchworm mechanism.`` In the manuscript, we made the effort to only use this term in reference to studies that were focused on this processes and described it as such -- namely refs 17 and 49 \noteg{update these after adding ref}. In our revisions, we have removed the term from the introduction so that it now reads as follows:

@@

\textcolor{blue}{10. I think some of the characterisation of the "all or nothing" picture is a bit misleading. The two state model of duplex formation is really a thermodynamic model, based on the assumption that the ensemble of hybridized and de-hybridized states look very different and are separated by a large barrier. This model is perfectly consistent with the transition between the two ensembles involving a complex network of metastable states. Indeed, as the authors note on line 3 of p4, a number of previous authors have considered what such states might look like, particularly for repetitive sequences. The authors' primary contribution here is not proposing that these states exist for the first time, but trying to tie together evidence from simulation and experiment to describe them.}

\noteg{The ``all-or-nothing'' comment is an important point. Brennan had cautioned using this term necessarily synonym for ``two-state'' and the reviewer seems to be making the same point. There are 11 instances of ``all-or-nothing'' in the main text, and it is often used in conjunction with ``two-state all-or-nothing.'' I think we may want to consider removing most of them, but we should probably word this carefully.}

\textbf{Author Reply:}   We appreciate the reviewer's helpful suggestion to highlight the role of macrostates rather than the identification of states themselves. We have updated some introductory language (in response comments from reviewer \#2 as well) to better communicate the novelty of the kinetic network:

\noteg{Same change as used above in response to \#2 wrt to better introductory language. Maybe some addtions in the main text or conclusions as well?}

### 2021-07-27 SI_marked_up.tex
just working on 3.7 and 3.8

\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

%\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
%\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{{\color{blue}{#1}}}
%\newcommand*{\blauwr}[1]{#1}

@@

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.70\textwidth]{FigS4.pdf}
    \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.}
    \label{fig:SIFig4}
\end{figure}

@@

@@

\begin{comment}

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Figs/figs_imp/GC-core_tracking_modes.pdf}
        \caption{Diffusion map representation configurations sampled from the F4 state. The x and y axes represent the the first two non-trivial diffusion map eigenfunctions. Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
        \label{fig:GC-core_dmaps}
	\end{center}
\end{figure}

\end{comment}

% shift to include this in the overall diffusion map section
%\subsubsection{Diffusion maps show ensemble of frayed states}

%We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate (Figure \ref{fig:GC-core_dmaps}). Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in ``overlap'' distance between adjacent A or T bases and the GC core. We defined overlap as the distance between a G/C and the A/T adjacent to its native complement. At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place. Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation. Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences. 
%\subsection{Explicit model comparisons}

%We also acknowledge that the implicit model may not fully capture the role of ions in facilitating hybridization and dissociation. In addition to implicit simulations, we followed the same pipeline to generate simulations and SRV-MSMs given the ion conditions specified in the \citet{Sanstead2016} using the 3SPN2 explicit implementation \citep{Hinckley2015}. These results look very similar to the implicit case, although with slightly slower dynamics and higher uncertainty in stationary probabilities. Comparisons were limited by the higher computational demand required to simulate 5x the beads in the box, a shorter integration timestep, and reliance on Ewald summation to account for electrostatics. We caution over-interpretation of these results as the explicit model has not been as widely adopted or thoroughly validated compared to the implicit model.

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_full.pdf}
%        \caption{Full 5$^\prime$ diffusion maps including degenerate third diffusion map mode.}
%        \label{fig:GC-end_dmaps_full}
%	\end{center}
%\end{figure}

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-end_dmaps_3prime.pdf}
%        \caption{Full 3$^\prime$ diffusion maps, shows greater overlap between the AT-all and GC-end populations, as well as a less distinct ``looping'' region for intact GC bonds}
%        \label{fig:GC-end_dmaps_3prime}
%	\end{center}
%\end{figure}

%add anymore figures for the time analysis: As shown bleow:
\begin{comment}

\subsubsection{\label{sec:Results}Code to provide on GitHub} 
\begin{enumerate}
    \item Cross val data
    \item lammps run script and each sequence input files
    \item Featurization with mdtraj (reindexing and sparse saving)
    \item Dmaps for GC-core and GC-end
    \item timescales comparison
    \item SRV cross valss
    \item MSM construction

\end{enumerate}
\end{comment}

\begin{comment}

\section{\label{sec:Discussion}Discussion}

\citep{Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}
Long simulations shows that terminal fraying does not converge on the picosecond timescale

\citep{Nonin1995TerminalFraying}
Most simulations suppress fraying by placing GC pairs at termini.

\citep{Andreatta2006UltrafastHelix}
- Base opening process is slower than 40 ns but faster than 1ms
- Open termini state shows new fraying mode of 5ps (I think shows free base motions)
\end{comment}

\begin{comment}

%%%%%% Other MSM and timescales approaches  %%%%%%%%%%%

McGibbon, R. T., & Pande, V. S. (2015). Variational cross-validation of slow dynamical modes in molecular kinetics : Importance of not generating too many microstates and employting cross-validation to avoid artificially high scores or artificially slow dynamics. Both of these lead to models fitting statistical noise which is the root of boosted scores.

Pinamonti, G., Paul, F., Rodriguez, A., & Bussi, G. (n.d.). analyzed with core-set Markov state models, 43: Good ideas on core-based clustering. Assigning microstates based on physical characteristics is interesting but didn't seem to work for me in practice (could try doing this based on sim.log data with number of base pairs bound or some energy metric). Lot's of notes on this paper regarding comparisons to timescales. Good notes on higher resolution fraying interactions that we likely can't resolve.

Sua, E., Adelman, J. L., & Zuckerman, D. M. (2016). Accurate Estimation of Protein Folding and Unfolding Times : Beyond Markov State Models, 1. https://doi.org/10.1021/acs.jctc.6b00339:
More accurate ways to adapt MFPT into a rate

Refs for DNA fraying:
\citep{Hagan2003AtomisticDNA} Atomistic simulations for biosensign app (can include in intro). Use TPS to focus on kinetic pathway by which singl base pair bind/unbind. Focus on end base of 3-bp oligomer in explicit solvent. Finds order parameters tracking base pair binding as well as intra-strand stacking. (energy and distance for each).

\citep{Wong2008TheSimulations} Three step melting process: untwisting most important,

6K.-Y. Wong and B. M. Pettitt, “The pathway of oligomeric DNA melting investigated by molecular dynamics simulations,” Biophys. J. 95, 5618– 5626 (2008).
37A. Perez and M. Orozco, “Real-time atomistic description of DNA unfold- ing,” Angew. Chem. 49, 4805–4808 (2010).
38M. F. Hagan, A. R. Dinner, D. Chandler, and A. K. Chakraborty, “Atomistic understanding of kinetic pathways for single base-pair binding and unbinding

\end{comment}

>>>>

\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand*{\rood}[1]{{\color{red}{#1}}}

%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
%\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{{\color{blue}{#1}}}
%\newcommand*{\blauwr}[1]{#1}

@@

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.6\textwidth]{FigS4.pdf}
    \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
    \label{fig:SIFig4}
\end{figure}

@@

\begin{figure}[ht!]
	\centering
    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS7.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First. each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
        %\label{fig:macrostate_colors_FES}
	\end{center}
\end{figure}

@@
